Novel lattice Boltzmann method for simulation of strongly shear thinning viscoelastic fluids

### 2024-03-05 NovelLBM.tex
L1 error definition

We start by simulating the alginate and \SI{0.49}{\percent} methyl cellulose solutions with a viscosity ratio $R_\eta = \num{48.2e3}$ and $R_\eta = \num{18.7}$, respectively, in 2D planar Poiseuille flow.
In physical units the channel has a height of \SI{20}{\micro\meter} which are discretized with 41 LBM nodes.
For the viscosity shuffling, we set $\alpha_\text{s} = 1$. The exact value of $\alpha_\text{s}$ does not matter as can be seen in figure~\ref{fig:ShuffleInPoiseuille}.
Along the flow direction, the channel is periodic, and the flow is driven by a pressure gradient of \SI{-1e7}{\pascal\per\meter} for alginate and  approximately \SI{-1.2e7}{\pascal\per\meter} for methyl cellulose.
The velocity field is initialized with zero.
For methyl cellulose, we simulate the full temporal evolution with the polymers starting in the unstretched configuration.
In the case of alginate, we speed up the convergence and start from a situation in which the polymers are pre-stretched. For this we calculate $\tau_{xx}$ and $\tau_{xy}$ using equations~(\ref{eq:tau_xx}) and (\ref{eq:PTTDerivation2}) using the $\dot{\gamma}$ calculated from equation~(\ref{eq:semiAnalyticGD}). The polymer stress is then initialized using these values. \SG{ give details }\RK{initialized to the semianalytical solution} \SG{ schreib am besten eine Referenz auf die Gleichung hin, muesste ja im Appendix stehen. }\RK{done}.
After \num{2e9} LBM time steps for alginate and \num{5e6} LBM time steps for methyl cellulose corresponding to approximately \SI{1.6}{\milli\second} and \SI{2.4}{\second} of physical time, respectively, the simulation is stopped and the flow field $u(y)$ is compared to the semi-analytical solution.
%\RK{
%	Alg 2e9 1.6455647956111251 ms (reveals the sim to be a scam)
%	MC 5e6 10.065719379854297 ms
%}
As shown in figure~\ref{fig:2Dvel}, very good agreement is found.
For a quantitative assessment, we compute the L2 error as follows.
\begin{equation}
	e_\text{L2} = \sqrt{\frac{\sum_i\left[\vec{u_i} - \vec{u}\left(y_i\right)\right]^2}{\sum_i\vec{u}\left(y_i\right)^2}}
\end{equation}
which gives \num{9e-4} and \num{1e-3} for alginate and methyl cellulose, respectively.
The spatially resolved errors are shown in figure~\ref{fig:2DvelErr} in the appendix.
These conclusions remain unchanged even when varying the pressure gradient, and thus the flow velocity, over many orders of magnitude as shown in figure~\ref{fig:pressureErr}~(a) of the appendix.

@@

We proceed to evaluate the viscosity shuffling method for a cylindrical Poiseuille flow.
The semi-analytical solution for this situation can be derived similarly as for the planar case and is given in appendix~\ref{app:semianalyticalPoiseuille}.
Numerically, the cylindrical situation is slightly more complex due to discretization errors ("staircase effect") when mapping the rounded channel walls onto the rectilinear LBM grid \cite{Lehmann.20229da}.
We use the same material parameters as above, but double the pressure gradient to \SI{-2e7}{\pascal\per\meter} and approximately \SI{-2.5e7}{\pascal\per\meter}, respectively, thereby mirroring the relation in Newtonian Poiseuille flow where twice the pressure is required to obtain the same flow rate in cylindrical compared to planar geometries.
The channel has a diameter of \SI{20}{\micro\meter} and is discretized with 41 lattice points.
Again, we initialize the polymers in the unstretched state for methyl cellulose and in the stretched state for alginate.
The resulting velocity profile can be seen in figure~\ref{fig:3Dvel} and shows almost equally good agreement as for the planar case with an L2 error of \num{0.014} and \num{0.06}, respectively.
The spatially resolved error in figure~\ref{fig:3DvelErr} of the appendix shows that this larger average is indeed mainly attributable to the staircase effect as the error near the walls increases significantly.
The robustness against pressure gradient changes is shown in figure~\ref{fig:pressureErr}~(b) of the appendix.

@@

%Malaspinas: We use richard/master fe0fc04115b9b6ea6628bcc6cb6593d3f72ac0ea (2023-11-05)
Finally, we assess the quality of our algorithm by comparing to an existing study on viscoelastic LBM by Malapsinas \textit{et al.} \cite{malaspinas2010}.
The authors studied an Oldroyd-B fluid in a planar Poiseuille flow and compared the error in the simulated velocity profile to an exact analytical solution.
The Oldroyd-B model is a special case of the PTT model which is obtained by setting $\xi = \epsilon = 0$ in eq.~(\ref{eq:polymerStress}) and is thus easily tractable with our code.\RK{was simulated with a direct Oldroyd-B implementation, which I now see was unnecessary...} \SG{ hast du mal kurz geprueft, ob dasselbe rauskommt? }\RK{Nein}\todo{check}
In figure~\ref{fig:malaspinas2010}, we show the L2 error as function of lattice resolution for different values of $\mathrm{Wi}$ and $R_ \nu $ which is defined as \cite{malaspinas2010}
\begin{align}
	R_\nu = \frac{\nu_\text{s}}{\nu_\text{p}+\nu_\text{s}}
\end{align}
with the kinematic solvent and the polymer viscosity $\nu_\text{s}$ and $\nu_\text{p}$. 
We note that this can directly be transformed into the viscosity ratio used above as $\frac{1}{R_\nu}-1=R_\eta$.
Furthermore, in agreement with \cite{malaspinas2010}, also the L2 error is defined differently than above, namely:
	\begin{equation}\label{eq:l2error}
		E_\text{u} = \frac{1}{u_\text{max}}\sqrt{\frac{1}{N}\sum\abs{u_\text{simulation}-u_\text{theory}}^2}.
	\end{equation}
At this point, we mention that there is a typographic mistake in eq.~(61) of \cite{malaspinas2010} where the prefactor should be 8 instead of 4.

@@

	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/2DAlginateErr.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/2DParameterStudyErrWorst.pdf}
		\caption{L2 error of the velocity profile in figure~\ref{fig:2Dvel} with respect to the semi-analytical theory in a planar Poiseuille flow for alginate (a) and methyl cellulose (b).
			The relative error is highest near the wall as the velocity is lowest there, but also because the simple half way bounce back boundaries used here allow some slip along the wall.}\RK{this is linear error} \SG{ define }
		\label{fig:2DvelErr}
	\end{figure}\todo{do}

	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/3DAlginateErr.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/3DParameterStudyErrWorst.pdf}
		\caption{L2 error of the velocity profile in \ref{fig:3Dvel} with respect to the semi-analytical theory in a cylindrical Poiseuille flow for alginate (a) and methyl cellulose (b).
			The relative error is highest near the wall as the velocity is lowest there, but also because the simple half way bounce back boundaries used here allow some slip along the wall.}\RK{this is linear error}
		\label{fig:3DvelErr}
	\end{figure}

>>>>

We start by simulating the alginate and \SI{0.49}{\percent} methyl cellulose solutions with a viscosity ratio $R_\eta = \num{48.2e3}$ and $R_\eta = \num{18.7}$, respectively, in 2D planar Poiseuille flow.
In physical units the channel has a height of \SI{20}{\micro\meter} which are discretized with 41 LBM nodes.
For the viscosity shuffling, we set $\alpha_\text{s} = 1$. The exact value of $\alpha_\text{s}$ does not matter as can be seen in figure~\ref{fig:ShuffleInPoiseuille}.
Along the flow direction, the channel is periodic, and the flow is driven by a pressure gradient of \SI{-1e7}{\pascal\per\meter} for alginate and  approximately \SI{-1.2e7}{\pascal\per\meter} for methyl cellulose.
The velocity field is initialized with zero.
For methyl cellulose, we simulate the full temporal evolution with the polymers starting in the unstretched configuration.
In the case of alginate, we speed up the convergence and start from a situation in which the polymers are pre-stretched. For this we calculate $\tau_{xx}$ and $\tau_{xy}$ using equations~(\ref{eq:tau_xx}) and (\ref{eq:PTTDerivation2}) using the $\dot{\gamma}$ calculated from equation~(\ref{eq:semiAnalyticGD}). The polymer stress is then initialized using these values. \SG{ give details }\RK{initialized to the semianalytical solution} \SG{ schreib am besten eine Referenz auf die Gleichung hin, muesste ja im Appendix stehen. }\RK{done}.
After \num{2e9} LBM time steps for alginate and \num{5e6} LBM time steps for methyl cellulose corresponding to approximately \SI{1.6}{\milli\second} and \SI{2.4}{\second} of physical time, respectively, the simulation is stopped and the flow field $u(y)$ is compared to the semi-analytical solution.
%\RK{
%	Alg 2e9 1.6455647956111251 ms (reveals the sim to be a scam)
%	MC 5e6 10.065719379854297 ms
%}
As shown in figure~\ref{fig:2Dvel}, very good agreement is found.
For a quantitative assessment, we compute the L2 error as follows.
\begin{equation}
	e_\text{L2} = \sqrt{\frac{\sum_i\left[\vec{u_i} - \vec{u}\left(y_i\right)\right]^2}{\sum_i\vec{u}\left(y_i\right)^2}}
\end{equation}
Using the simulated velocities $\vec{u_i}$ and the semi-analytical results $\vec{u}$. This gives \num{9e-4} and \num{1e-3} for alginate and methyl cellulose, respectively.
The spatially resolved L1 error defined as follows is shown in figure~\ref{fig:2DvelErr} in the appendix.
\begin{equation}
	e_\text{L1} = \frac{\vec{u_i} - \vec{u}\left(y_i\right)}{\vec{u}\left(y_i\right)}
\end{equation}
\RK{This error definition is new}
These conclusions remain unchanged even when varying the pressure gradient, and thus the flow velocity, over many orders of magnitude as shown in figure~\ref{fig:pressureErr}~(a) of the appendix.

@@

We proceed to evaluate the viscosity shuffling method for a cylindrical Poiseuille flow.
The semi-analytical solution for this situation can be derived similarly as for the planar case and is given in appendix~\ref{app:semianalyticalPoiseuille}.
Numerically, the cylindrical situation is slightly more complex due to discretization errors ("staircase effect") when mapping the rounded channel walls onto the rectilinear LBM grid \cite{Lehmann.20229da}.
We use the same material parameters as above, but double the pressure gradient to \SI{-2e7}{\pascal\per\meter} and approximately \SI{-2.5e7}{\pascal\per\meter}, respectively, thereby mirroring the relation in Newtonian Poiseuille flow where twice the pressure is required to obtain the same flow rate in cylindrical compared to planar geometries.
The channel has a diameter of \SI{20}{\micro\meter} and is discretized with 41 lattice points.
Again, we initialize the polymers in the unstretched state for methyl cellulose and in the stretched state for alginate.
The resulting velocity profile can be seen in figure~\ref{fig:3Dvel} and shows almost equally good agreement as for the planar case with an L2 error of \num{0.014} and \num{0.06}, respectively.
The spatially resolved L1 error in figure~\ref{fig:3DvelErr} of the appendix shows that this larger average is indeed mainly attributable to the staircase effect as the error near the walls increases significantly.
The robustness against pressure gradient changes is shown in figure~\ref{fig:pressureErr}~(b) of the appendix.

@@

%Malaspinas: We use richard/master fe0fc04115b9b6ea6628bcc6cb6593d3f72ac0ea (2023-11-05)
Finally, we assess the quality of our algorithm by comparing to an existing study on viscoelastic LBM by Malapsinas \textit{et al.} \cite{malaspinas2010}.
The authors studied an Oldroyd-B fluid in a planar Poiseuille flow and compared the error in the simulated velocity profile to an exact analytical solution.
The Oldroyd-B model is a special case of the PTT model which is obtained by setting $\xi = \epsilon = 0$ in eq.~(\ref{eq:polymerStress}) and is thus easily tractable with our code.\RK{was simulated with a direct Oldroyd-B implementation, which I now see was unnecessary...} \SG{ hast du mal kurz geprueft, ob dasselbe rauskommt? }\RK{Ja, ist aber auch von code offensichtlich, der unterschied wird zur OpenCL compile time entfernt}
In figure~\ref{fig:malaspinas2010}, we show the L2 error as function of lattice resolution for different values of $\mathrm{Wi}$ and $R_ \nu $ which is defined as \cite{malaspinas2010}
\begin{align}
	R_\nu = \frac{\nu_\text{s}}{\nu_\text{p}+\nu_\text{s}}
\end{align}
with the kinematic solvent and the polymer viscosity $\nu_\text{s}$ and $\nu_\text{p}$. 
We note that this can directly be transformed into the viscosity ratio used above as $\frac{1}{R_\nu}-1=R_\eta$.
Furthermore, in agreement with \cite{malaspinas2010}, also the L2 error is defined differently than above, namely:
	\begin{equation}\label{eq:l2error}
		E_\text{u} = \frac{1}{u_\text{max}}\sqrt{\frac{1}{N}\sum\abs{u_\text{simulation}-u_\text{theory}}^2}.
	\end{equation}
At this point, we mention that there is a typographic mistake in eq.~(61) of \cite{malaspinas2010} where the prefactor should be 8 instead of 4.

@@

	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/2DAlginateErr.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/2DParameterStudyErrWorst.pdf}
		\caption{L1 error of the velocity profile in figure~\ref{fig:2Dvel} with respect to the semi-analytical theory in a planar Poiseuille flow for alginate (a) and methyl cellulose (b).
			The relative error is highest near the wall as the velocity is lowest there, but also because the simple half way bounce back boundaries used here allow some slip along the wall.}\RK{this is linear error} \SG{ define }\RK{done}
		\label{fig:2DvelErr}
	\end{figure}

	\begin{figure}[H]
		\centering
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/3DAlginateErr.pdf}
		\adjincludegraphics[trim=0 0 0 0, clip, scale=1]{assets/plots/3DParameterStudyErrWorst.pdf}
		\caption{L1 error of the velocity profile in \ref{fig:3Dvel} with respect to the semi-analytical theory in a cylindrical Poiseuille flow for alginate (a) and methyl cellulose (b).
			The relative error is highest near the wall as the velocity is lowest there, but also because the simple half way bounce back boundaries used here allow some slip along the wall.}
		\label{fig:3DvelErr}
	\end{figure}
